Modeling Interactions of Ions with Ether- and Ester-linked Phospholipids

### 2019-11-20 paper.tex
Final Commit

\documentclass[12pt,openany,final]{book}
\usepackage[letterpaper,margin=1in,footskip=0.25in]{geometry}
\fontfamily{cmr}
\usepackage[final]{changes}
\usepackage{lipsum}
\usepackage[nottoc]{tocbibind}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{flafter}
\usepackage{floatrow}
%\usepackage{floatflt}
\usepackage{placeins}
\usepackage{siunitx}
\usepackage{graphicx}
\graphicspath{{figures/}}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{appendix}
\usepackage{amsmath}
\usepackage{cases}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{color}
\usepackage{enumitem}
\usepackage{soul}
\usepackage{indentfirst}
\usepackage{hyphenat}
\usepackage{xspace}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{xcolor}
\usepackage{lineno}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{hyperref}
\usepackage[superscript]{cite}
\usepackage{lineno}
\titleformat
{\chapter} % command
[display] % shape
{\centering\sc\LARGE} % format
{\centering Chapter \thechapter} % label
{0.2ex} % sep
{
} % before-code
\titlespacing*{\chapter}{0pt}{1in}{0.25in}
\titlespacing*{\section}{0pt}{0pt}{0pt}
\hbadness=99999
%\epstopdfDeclareGraphicsRule{.tga}{png}{.png}{
%    convert #1 \OutputFile
%}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}
%\newcommand{\cmpersec}{$\frac{\text{cm}}{\text{s}}$~}
\newcommand{\cmpersec}{cm/s~}
\newcommand{\etal}{\textit{et al.}}
\newcommand{\na}{Na\textsuperscript{+}~}
\newcommand{\cl}{Cl\textsuperscript{-}~}
\newcommand{\PM}{$\pm$~}
\newcommand{\invnm}{$\text{\si{\nm}}^{-1}$~}
\newcommand{\nmeter}{\si{\nm}~}
\renewcommand{\arraystretch}{1.8}
%\newcommand{\hl}{\added}
%\newcommand{\sout}{\deleted}
\newcommand{\TODO}{\hl{TODO}~}
\newcommand{\about}{$\sim$}
\newcommand{\aangstroms}{{\aa}ngstr{\"o}ms}
\newcommand{\aangstrom}{{\aa}ngstr{\"o}m}
\newenvironment{hle}{\color{red}}

\newcommand{\beginsupplemental}{%
    \setcounter{table}{0}
    \renewcommand{\thetable}{S\arabic{table}}%
    \setcounter{figure}{0}
    \renewcommand{\thefigure}{S\arabic{figure}}%
}

\author{Matthew Saunders}
\title{Modeling Interactions of Ions with Ether- and Ester-linked Phospholipids}
\begin{document}
\pagestyle{plain}
%\maketitle
\begin{titlepage}
\begin{centering}
~\\[1in]
Modeling of Interaction of Ions \\[2\baselineskip] with Ether- and Ester-linked phospholipids
~\\[3\baselineskip]
by
~\\[3\baselineskip]
Matthew W. Saunders
~\\[4\baselineskip]
A thesis submitted in partial fufillment\\
of the requirements of the degree of\\
Master of Science in Biology \\ with a concentration in Cell \& Molecular Biology\\
Department of Cell Biology, Microbiology and Molecular Biology\\
College of Arts and Sciences\\
University of South Florida\\
~\\[2\baselineskip]
Major Professor: Sameer Varma, Ph.D.\\
Sagar Pandit, Ph.D.\\
H. Lee Woodcock, Ph.D.\\
Jianjun Pan, Ph.D.\\
~\\[2\baselineskip]
%Date of Approval:\\
%XXX\\
~\\[3\baselineskip]
\end{centering}
\end{titlepage}
\doublespacing
\tableofcontents
\pagenumbering{roman}
\listoftables
\listoffigures
\bibliographystyle{elsarticle-num}
%\linenumbers
\doublespacing
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Polar lipids, or phospholipids, are present in all parts of cells
and are used in many signalling and structural roles.
As structural molecules they act as the main component of cellular plasma membranes --- bilayers
of polar lipids that separate cells from their environments. Bilayer properties are heavily
influenced by the structure of their component polar lipids, and different lipids
are found in different organisms. 
A distinguishing feature of Archaeal plasma membranes is that their phospholipids contain ether-links, as opposed to bacterial and eukaryotic plasma
membranes where phospholipids primarily contain ester-links. In our work we examine the effects of
salt on bilayer structure in the case of an ether-linked lipid bilayer. We use molecular dynamics simulations and compare equilibrium properties of two model lipid
bilayers in NaCl salt solution --- POPC and its ether-linked analog that we refer to as HOPC. We make the following key observations.
The headgroup region of HOPC~``adsorbs'' fewer ions compared to the headgroup region of POPC. Consistent with this,
we note that the Debye screening length in the HOPC system is \about 10\% shorter than that
in the POPC system. Herein, we introduce a protocol to identify the lipid-water interfacial boundary that
reproduces the bulk salt distribution consistent with
Gouy-Chapman theory. We also note that the HOPC bilayer has excess solvent
in the headgroup region when compared to POPC,
coinciding with a trough in the electrostatic potential. Waters in
this region have longer autocorrelation times and smaller lateral diffusion rates compared to the corresponding region
in the POPC bilayer, suggesting that the waters in HOPC are more strongly
coordinated to the lipid headgroups. Furthermore, we note that it is this
region of tightly coordinated waters in the HOPC system that has a lower density of Na$^+$ ions.
Based on these observations we conclude that an ether-linked lipid bilayer has a lower binding
affinity for Na$^+$ compared to an ester-linked lipid bilayer.

\chapter{Introduction}
\linenumbers
\pagenumbering{arabic}

\section{Ether- and Ester-linked Phospholipids}
Phospholipids are a diverse species of biological molecules used in many roles in life~\cite{van:2008:lipidvariety}. These molecules
are amphipathic, with a headgroup that is hydrophillic attached to hydrophobic acyl chains by a
glycerol backbone~\cite{pandit:2008:simulationtextbook,israelachvili:2011:intermol}. 
These molecules, when above a critical concentration in aqueous solution, self assemble into
various aggregate structures in order 
to reduce the exposure of the hydrophobic chains
to polar solvent --- these structures include
micelles, vesicles, and bilayers along with others~\cite{israelachvili:2011:intermol}. 
Phospholipids tend to form bilayers, 
where the hydrophillic headgroups face the solvent and the 
hydrophobic acyl chains face into the center 
of the structure~\cite{israelachvili:2011:intermol,ashrafuzzaman:2012:membrane}. 
An image of a bilayer of a commonly studied
model lipid can be seen
in figure~\ref{fig:boximage}.
\begin{figure}[htp]
    \caption[Image of a model bilayer of phospholipids from a simulation.]{ Image of a model bilayer of phospholipids from a simulation. The lipids can be seen
in the center of the image, with the hydrophobic chains colored in
light purple. The polar headgroup can be seen highlighted by the phosphate group in orange and the choline nitrogen in blue
facing the solvent. Solvent molecules have been removed for clarity. Figure generated
using VMD~\cite{vmd}}
\label{fig:boximage}
\includegraphics[height=0.25\textheight]{images/popc_image}
\end{figure}
In living cells, lipid bilayers act as the delimiter between the cell and the environment, acting as one of the major constituents of the cellular 
plasma membrane~\cite{reece:2014:biobook}. The plasma membrane consists of a mixture of many lipid species, as
well as proteins and lipid-like molecules, and the mixture is tuned by the cell to adapt the 
structure and flexibility of the plasma membrane to different environments~\cite{pandit:2008:simulationtextbook,van:2008:lipidvariety,zhang:2008:memhomeo}. 
Furthermore, the cellular membrane is the interface for exchange
of nutrients with the environment, and also the interface for the transmission of signals to and from the outside of the cell. This
exchange is done through passive diffusion, proteins and other molecules embedded into the membrane, and through the formation
of vesicles and buds out of the lipid bilayer~\cite{reece:2014:biobook}. The behavior and function of the plasma membrane
is heavily dependent on the structure of the bilayer, which again can be influenced directly by the mixture of phospholipids constituting the
membrane~\cite{van:2008:lipidvariety,zhang:2008:memhomeo}.

@@

Phospholipids can also use several types of chemical links between the glycerol
backbone and the fatty acid chains~\cite{sehgal:1962, koga:2014}.
Phospholipids containing ether-links represent one of the many major evolutionary divides -- 
while they make up much of the phospholipids of archaeal membranes,
 they are present in only small fractions in eukarya and bacteria~\cite{sehgal:1962, koga:2014, kates:1993:archealipids}. 
These phospholipids differ from the more common ester-linked phospholipids of bacterial and 
eukaryotic membranes in that they lack fatty acid carbonyl groups (see figure \ref{fig:struc}).

\begin{figure}[htp]
\caption[Chemical structures of representative phospholipids highlighting the difference
between ether- and ester-links.]{ 
Chemical structures of representative phospholipids highlighting the difference 
between ether- and ester-links. POPC (1-palmitoyl-2-oleoyl-sn-glycero-3-phosphatidylcholine) is a 
diester lipid and HOPC (1-hexadecyl-2-(9-octadecenyl)-sn-glycero-3-phosphatidylcholine) is its diether analog that lacks fatty acid carbonyl groups.
}
\label{fig:struc}
\includegraphics[height=0.5\textheight]{images/Structure_annotated}
\end{figure}

@@

Nevertheless, all studies aimed at examining differences between ester- and ether-linked lipid bilayers have been carried out in pure water, except for the one recent study by 
Leonard \etal~\cite{leonard:2018}. As such, several studies show that salt significantly 
affects the properties of ester-linked lipid bilayers~\cite{kruczek:2019,kruczek:2017, duro:2016,pabst:2007,sachs:2004,petrache:2006:swelling}. 
For example, in our previous works on the interactions of monovalent and divalent ions on zwitterionic ester-linked lipid bilayers~\cite{kruczek:2019,kruczek:2017}
we found that monovalent cations insert into headgroup region. 
We reported reduction in area per lipid, increase in  bilayer thickness, increased range of ordered solvent at the bilayer surface, and an increase in
the electrostatic dipole potential of the bilayer.
Despite the work done on these lipids in the past, the question of how ions interact with bilayers of ether-linked
phospholipids compared to ester-linked remains a topic of interest.

@@

The earliest lipid simulations often used a united atom representation of the lipids, which simplified lipid chains by expressing carbon groups
as beads rather than explicitly defining hydrogens~\cite{raghavan:1992,marrink:1993,egberts:1994}. These simulations focused on validation with experimental
values for bilayer structure, such as area per lipid and bilayer thickness that are taken from SAXS and SANS measurements. 
Later development of force-fields included calculating partial charges
for different parts of the lipid molecules using \emph{ab initio} quantum calculations~\cite{chiu:1995:incorporation,marrink:1993}. Chiu \etal
calculated charges in this way for the GROMOS force-field of DPPC in 1995~\cite{chiu:1995:incorporation}.

Berger \etal sought to further improve the densities given by the GROMOS force-field in 1997 
by optimizing VdW parameters using molecular volumes and
heats of vaporization in small hydrocarbon molecules to improve the acyl chains of DPPC~\cite{berger:1997},
work that was continued by Chiu \etal~\cite{chiu:1999:optimization,chiu:2003:structure}, 
Later, further optimization was carried out by Chiu \etal 
in 2009 to improve torsional and VdW parameters using experimental 
heats of vaporization and densities, and validated
the overall result by predicting small-angle x-ray scattering 
results for model lipid bilayers to create the GROMOS 43-A1S3 lipid force field which gave excellent reproduction of SAXS form-factors for
different lipids~\cite{chiu:2009}.

Molecular dynamics can give this resolution while still reproducing the large-scale properties of the system as a whole~\cite{pandit:2008:simulationtextbook,dror:2012:biomicroscope}. 
Properties of the lipid bilayer can be expressed directly in terms of the positions or dynamics of atoms, 
providing additional understanding of these systems.

@@

\chapter{Interaction of salt with ether- and ester-linked phospholipid bilayers}
In this work we seek to answer the general questions of how salt inte\-racts 
with ether-linked lipid bilayers, and how ether-linked lipid bilayers differ from ester-linked lipid bilayers in the presence of salt.
We addressed these questions by carrying out a molecular dynamics simulation of 
1-hexadecyl-2-(9-octadecenyl)-sn-glycero-3-phosphatidylcholine lipid bilayer (HOPC) in NaCl salt. 
We also compared results of this simulation against our previous simulation of 
1-palmitoyl-2-oleoyl-sn-glycero-3-phosphatidylcholine lipid bilayer (POPC) 
in NaCl \cite{kruczek:2017}. Additionally, to gain insight into how salt modulates differences between ether- 
and ester-linked lipid bilayers, we also compared these simulations against those of ether- and ester-linked lipid bilayers in pure water~\cite{kruczek:2017:ether}. 
\section{Methods}
The trajectories for the POPC bilayer in 200 mM NaCl were taken from our previous work~\cite{kruczek:2017}, 
and were re-analyzed to address the specific goals of this project. The trajectories
for the HOPC bilayer were generated as described below, following the same protocol we used for the POPC bilayer. 
Analysis was carried out using GROMACS~\cite{abraham:2015,pall:2014,van:2005,lindahl:2001,berendsen:1995} utility 
tools, and in-house software developed using GROMACS API.

@@

\begin{figure}[p]
    \caption[Comparison of component mass densities between the HOPC and POPC systems.]{ 
Comparison of component mass densities between the HOPC and POPC systems. 
Densities are computed by dividing the box dimension along the membrane normal into 2000 slices. For 
clairity of presentation, error bars are indicated for every 10th slice. 
Mass densities of solvent and ions, water, lipid headgroup and backbone, and lipid chains are shown. 
The plot for HOPC shows a peak in the water density inside the headgroup region that is not present in POPC, 
similar to what was seen in our previous work with DHPC~\cite{kruczek:2017:ether}. 
}
\label{fig:massdens}
\includegraphics[width=	0.7\textwidth]{dens_3component.eps}
\end{figure}
We find that the headgroup region of 
HOPC has a higher density of solvent compared to POPC -- this is visible as 
a distinct peak in the solvent density profile in the HOPC system. 
We verified that this additional peak does not
result from accumulation of ions by comparing mass density profiles of only water molecules. This
peak is similar to observations of a smaller peak 
made in our previous simulations in the absence of salt,
and may be a direct result of the ether-linked lipids~\cite{kruczek:2017:ether}.

>>>>

\documentclass[12pt,openany,final]{book}
\usepackage[letterpaper,margin=1in,footskip=0.25in]{geometry}
\fontfamily{cmr}
\usepackage[final]{changes}
\usepackage{lipsum}
\usepackage[nottoc]{tocbibind}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{flafter}
\usepackage{floatrow}
\usepackage{float}
\floatstyle{plaintop}
\restylefloat{table}
%\usepackage{floatflt}

\usepackage{placeins}
\usepackage{siunitx}
\usepackage{graphicx}
\graphicspath{{figures/}}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{appendix}
\usepackage{amsmath}
\usepackage{cases}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{color}
\usepackage{enumitem}
\usepackage{soul}
\usepackage{indentfirst}
\usepackage{hyphenat}
\usepackage{xspace}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{xcolor}
\usepackage{lineno}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{hyperref}
\usepackage[superscript]{cite}
\usepackage{lineno}
\titleformat
{\chapter} % command
[display] % shape
{\centering\sc\LARGE} % format
{\centering Chapter \thechapter} % label
{0.2ex} % sep
{
} % before-code
\titlespacing*{\chapter}{0pt}{0.5in}{0.25in}
\titlespacing*{\section}{0pt}{0pt}{0pt}
\hbadness=99999
%\epstopdfDeclareGraphicsRule{.tga}{png}{.png}{
%    convert #1 \OutputFile
%}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}
\usepackage[titles]{tocloft}
\setlength\cftbeforechapskip{\baselineskip}
\setlength\cftbeforefigskip{\baselineskip}
\setlength\cftbeforetabskip{\baselineskip}
\setlength\cftfigindent{0in}
\setlength\cfttabindent{0in}
\makeatletter
\setlength{\@fptop}{0pt}
\makeatother
\usepackage{url}
\urlstyle{same}

%\newcommand{\cmpersec}{$\frac{\text{cm}}{\text{s}}$~}
\newcommand{\cmpersec}{cm/s~}
\newcommand{\etal}{\textit{et al.}}
\newcommand{\na}{Na\textsuperscript{+}~}
\newcommand{\cl}{Cl\textsuperscript{-}~}
\newcommand{\PM}{$\pm$~}
\newcommand{\invnm}{$\text{\si{\nm}}^{-1}$~}
\newcommand{\nmeter}{\si{\nm}~}
\renewcommand{\arraystretch}{1.8}
%\newcommand{\hl}{\added}
%\newcommand{\sout}{\deleted}
\newcommand{\TODO}{\hl{TODO}~}
\newcommand{\about}{$\sim$}
\newcommand{\aangstroms}{{\aa}ngstr{\"o}ms}
\newcommand{\aangstrom}{{\aa}ngstr{\"o}m}
\newenvironment{hle}{\color{red}}

\newcommand{\beginsupplemental}{%
    \setcounter{table}{0}
    \renewcommand{\thetable}{S\arabic{table}}%
    \setcounter{figure}{0}
    \renewcommand{\thefigure}{S\arabic{figure}}%
}
\renewcommand{\bibname}{References}
\titleformat{\chapter}
{\normalfont\normalsize\bfseries\centering}{\thechapter}{1em}{}
\titleformat{\section}
{\normalfont\normalsize\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}
{\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
\titleformat{\subsubsection}
{\normalfont\normalsize\bfseries}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]
{\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
\titleformat{\subparagraph}[runin]
{\normalfont\normalsize\bfseries}{\thesubparagraph}{1em}{}
%\cftsetpnumwidth{0.5in}
%\setlength{\cftbeforefigskip}{-\baselineskip}
%\setlength{\cftbeforetabskip}{-\baselineskip}
\cftsetpnumwidth{0.5in}
\cftsetrmarg{0.5in}
\renewcommand{\cftdot}{}
\renewcommand{\cftchapfont}{\normalsize}
\renewcommand{\cftchappagefont}{\normalsize}

\author{Matthew Saunders}
\title{Modeling Interactions of Ions with Ether- and Ester-linked Phospholipids}
\begin{document}

\pagestyle{plain}
%\maketitle
\begin{titlepage}
\begin{centering}

~\\[1in]
Modeling of Interaction of Ions with Ether- and Ester-linked phospholipids
~\\[3\baselineskip]
by
~\\[3\baselineskip]
Matthew W. Saunders
~\\[4\baselineskip]
A thesis submitted in partial fufillment\\
of the requirements for the degree of\\
Master of Science in Biology \\ with a concentration in Cell \& Molecular Biology\\
Department of Cell Biology, Microbiology and Molecular Biology\\
College of Arts and Sciences\\
University of South Florida
~\\[2\baselineskip]
Co-Major Professor: Sameer Varma, Ph.D.\\
Co-Major Professor: Sagar Pandit, Ph.D.\\
H. Lee Woodcock, Ph.D.\\
Jianjun Pan, Ph.D.
~\\[2\baselineskip]
Date of Approval:~\\
October 23\textsuperscript{rd}, 2019
~\\[3\baselineskip]
Keywords: Molecular Dynamics, Gouy-Chapman Theory\\
~\\
Copyright \textsuperscript\textcopyright~2019, Matthew W. Saunders\\

\end{centering}
\end{titlepage}
\tableofcontents
\pagenumbering{roman}
\listoftables
\listoffigures
\bibliographystyle{elsarticle-num}
%\linenumbers
\doublespacing
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Phospholipids are present in all parts of cells
and are used in many signalling and structural roles.
As structural molecules they act as the main component of cellular membranes. 
Bilayer properties are heavily
influenced by the structure of their component polar lipids, and different lipids
are found in different organisms. 
A distinguishing feature of Archaeal plasma membranes is that their phospholipids contain ether-links, as opposed to bacterial and eukaryotic plasma
membranes where phospholipids primarily contain ester-links. In our work we examine the effects of
salt on bilayer structure in the case of both ester- and ether-linked lipid bilayers. 
We use molecular dynamics simulations and compare equilibrium properties of two model lipid
bilayers in NaCl salt solution --- POPC and its ether-linked analog that we refer to as HOPC. We make the following key observations.
The headgroup region of HOPC~``adsorbs'' fewer ions compared to the headgroup region of POPC. Consistent with this,
we note that the Debye screening length in the HOPC system is \about 10\% shorter than that
in the POPC system. Herein, we introduce a protocol to identify the lipid-water interfacial boundary that
reproduces the bulk salt distribution consistent with
Gouy-Chapman theory. We also note that the HOPC bilayer has excess solvent
in the headgroup region when compared to POPC,
coinciding with a trough in the electrostatic potential. Waters in
this region have longer autocorrelation times and smaller lateral diffusion rates compared to the corresponding region
in the POPC bilayer, suggesting that the waters in HOPC are more strongly
coordinated to the lipid headgroups. Furthermore, we note that it is this
region of tightly coordinated waters in the HOPC system that has a lower density of Na$^+$ ions.
Based on these observations we conclude that an ether-linked lipid bilayer has a lower binding
affinity for Na$^+$ compared to an ester-linked lipid bilayer.

\chapter[Introduction]{Introduction\footnote{
    Portions of this chapter have been previously published in \emph{Biochimica et Biophysica Acta (BBA) - Biomembranes} Volume 1861, Issue 5, 1 May 2019, Pages 907-915,
    and have been reproduced with permission from Elsevier\cite{saunders:2019}.
}}
%\linenumbers
\pagenumbering{arabic}
\section{Ether- and Ester-linked Phospholipids}
Phospholipids are a diverse species of biological molecules used in many roles in life~\cite{van:2008:lipidvariety}. These molecules
are amphipathic, with a headgroup that is hydrophillic attached to hydrophobic acyl chains by a
glycerol backbone~\cite{pandit:2008:simulationtextbook,israelachvili:2011:intermol}. 
These molecules, when above a critical concentration in aqueous solution, self assemble into
various aggregate structures in order 
to reduce the exposure of the hydrophobic chains
to polar solvent --- these structures include
micelles, vesicles, and bilayers along with others~\cite{israelachvili:2011:intermol}. 
Phospholipids tend to form bilayers, 
where the hydrophillic headgroups face the solvent and the 
hydrophobic acyl chains face into the center 
of the structure~\cite{israelachvili:2011:intermol,ashrafuzzaman:2012:membrane}. 
%An image of a bilayer of a commonly studied
%model lipid can be seen
%in figure~\ref{fig:boximage}.
%\begin{figure}[htp]
%    \caption[Image of a model bilayer of phospholipids from a simulation.]{ Image of a model bilayer of phospholipids from a simulation. The lipids can be seen
%in the center of the image, with the hydrophobic chains colored in
%light purple. The polar headgroup can be seen highlighted by the phosphate group in orange and the choline nitrogen in blue
%facing the solvent. Solvent molecules have been removed for clarity. Figure generated
%using VMD~\cite{vmd}}
%\label{fig:boximage}
%\includegraphics[height=0.25\textheight]{images/popc_image}
%\end{figure}
In living cells, lipid bilayers help to delimit the cell and the environment, acting as one of the major constituents of the cellular 
plasma membrane~\cite{reece:2014:biobook}. The plasma membrane consists of a mixture of many lipid species, as
well as proteins and lipid-like molecules, and the mixture is tuned by the cell to adapt the 
structure and flexibility of the plasma membrane to different environments~\cite{pandit:2008:simulationtextbook,van:2008:lipidvariety,zhang:2008:memhomeo}. 
Furthermore, the cellular membrane is the interface for exchange
of nutrients with the environment, and also the interface for the transmission of signals to and from the outside of the cell. This
exchange is done through passive diffusion, proteins, and other molecules embedded into the membrane, and through the formation
of vesicles and buds out of the lipid bilayer~\cite{reece:2014:biobook}. The behavior and function of the plasma membrane
is heavily dependent on the structure of the bilayer, which again can be influenced directly by the mixture of phospholipids constituting the
membrane~\cite{van:2008:lipidvariety,zhang:2008:memhomeo}.

@@

Phospholipids can also use several types of chemical links between the glycerol
backbone and the fatty acid chains~\cite{sehgal:1962, koga:2014}.
Phospholipids containing ether-links represent one of the many major evolutionary divides --- 
while they make up much of the phospholipids of archaeal membranes,
 they are present in only small fractions in eukarya and bacteria~\cite{sehgal:1962, koga:2014, kates:1993:archealipids}. 
These phospholipids differ from the more common ester-linked phospholipids of bacterial and 
eukaryotic membranes in that they lack fatty acid carbonyl groups (see figure \ref{fig:struc}).

\begin{figure}[htp]
\caption[Chemical structures of representative phospholipids highlighting the difference
between ether- and ester-links.]{ 
Chemical structures of representative phospholipids highlighting the difference 
between ether- and ester-links. POPC (1-palmitoyl-2-oleoyl-sn-glycero-3-phosphatidylcholine) is a 
diester lipid and HOPC (1-hexadecyl-2-(9-octadecenyl)-sn-glycero-3-phosphatidylcholine) is its diether analog that lacks fatty acid carbonyl groups.
}
\label{fig:struc}
\includegraphics[height=0.8\textheight]{images/Structure_annotated}
\end{figure}

@@

Nevertheless, all studies aimed at examining differences between ester- and ether-linked lipid bilayers have been carried out in pure water, except for the one recent study by 
Leonard \etal where an all-atom force field was parameterized for a different model ether-linked lipid~\cite{leonard:2018}. In the work by Leonard \etal, the primary analysis
given was for systems without salt, and did not treat the system with salt beyond structural measurements~\cite{leonard:2018}. The questions of how ions bind differenty to 
bilayers of ether- and ester-linked lipids as well as how they interact with the bilayers electrostatically are still open to investigation. The inclusion of
ions in a bilayer system is an important step to acheiving a more phisiologically relevant system --- biological membranes are always exposed to ions~\cite{reece:2014:biobook}. 
In biology, the affects of ions on bilayer structure cannot be ignored.
%TODO Keep adding things here I think it needs more %%
As such, several studies show that salt significantly 
affects the properties of ester-linked lipid bilayers~\cite{kruczek:2019,kruczek:2017,duro:2016,pabst:2007,sachs:2004,petrache:2006:swelling,leonard:2018}. 
For example, in our previous works on the interactions of monovalent and divalent ions on zwitterionic ester-linked lipid bilayers~\cite{kruczek:2019,kruczek:2017}
we found that monovalent cations insert into headgroup region. 
We reported reduction in area per lipid, increase in  bilayer thickness, increased range of ordered solvent at the bilayer surface, and an increase in
the electrostatic dipole potential of the bilayer.
Despite the work done on these lipids in the past, the question of how ions interact with bilayers of ether-linked
phospholipids compared to ester-linked remains a topic of interest.

@@

The earliest lipid simulations often used a united atom representation of the lipids, which simplified lipid chains by expressing carbon groups
as beads rather than explicitly defining hydrogens~\cite{raghavan:1992,marrink:1993,egberts:1994}. These simulations focused on validation with experimental
values for bilayer structure, such as area per lipid and bilayer thickness that are taken from SAXS and SANS measurements. 
Later development of force-fields included calculating partial charges
for different parts of the lipid molecules using \emph{ab initio} quantum calculations~\cite{chiu:1995:incorporation,marrink:1993}. Chiu \etal
calculated charges in this way for the GROMOS 43-A1 force-field of DPPC in 1995~\cite{chiu:1995:incorporation}.

Berger \etal sought to further improve the densities given by the GROMOS force-field in 1997 
by optimizing VdW parameters using molecular volumes and
heats of vaporization in small hydrocarbon molecules to improve the acyl chains of DPPC~\cite{berger:1997},
work that was continued by Chiu \etal~\cite{chiu:1999:optimization,chiu:2003:structure}, 
Later, further optimization was carried out by Chiu \etal 
in 2009 to improve torsional and VdW parameters using experimental 
heats of vaporization and densities, and validated
the overall result by predicting small-angle x-ray scattering 
results for model lipid bilayers to create the GROMOS 43-A1S3 lipid force field that we use for this work, which gave excellent reproduction of SAXS form-factors for
different lipids~\cite{chiu:2009}. Parameters for the ether-links were added to the force-field by Kruczek \etal in 2017~\cite{kruczek:2017}, and the force-field
was demonstrated to work well with ion parameters from Joung and Cheatham III~\cite{joung:2008} by reproucing ion binding constants in the same
year~\cite{kruczek:2017}.

%Molecular dynamics can give this resolution while still reproducing the large-scale properties of the system as a whole~\cite{pandit:2008:simulationtextbook,dror:2012:biomicroscope}. 
%Properties of the lipid bilayer can be expressed directly in terms of the positions or dynamics of atoms, 
%providing additional understanding of these systems.

@@

\chapter[Interaction of salt with ether- and ester-linked phospholipid bilayers]{Interaction of salt with ether- and ester-linked phospholipid bilayers\footnote{
    Portions of this chapter have been previously published in \emph{Biochimica et Biophysica Acta (BBA) - Biomembranes} Volume 1861, Issue 5, 1 May 2019, Pages 907-915,
    and have been reproduced with permission from Elsevier\cite{saunders:2019}.
}}
In this work we seek to answer the general questions of how salt inte\-racts 
with ether-linked lipid bilayers, and how ether-linked lipid bilayers differ from ester-linked lipid bilayers in the presence of salt.
We addressed these questions by carrying out a molecular dynamics simulation of 
1-hexadecyl-2-(9-octadecenyl)-sn-glycero-3-phosphatidylcholine lipid bilayer (HOPC) in NaCl salt. 
We also compared results of this simulation against our previous simulation of 
1-palmitoyl-2-oleoyl-sn-glycero-3-phosphatidylcholine lipid bilayer (POPC) 
in NaCl \cite{kruczek:2017}. Additionally, to gain insight into how salt modulates differences between ether- 
and ester-linked lipid bilayers, we also compared these simulations against those of ether- and ester-linked lipid bilayers in pure water~\cite{kruczek:2017:ether}. 
\section{Methods}
The trajectories for the POPC bilayer in 200 mM NaCl were taken from our previous work~\cite{kruczek:2017}, 
and were re-analyzed to address the specific goals of this project. The trajectories
for the HOPC bilayer were generated as described below, following the same protocol we used for the POPC bilayer. 
Analysis was carried out using GROMACS~\cite{abraham:2015,pall:2014,van:2005,lindahl:2001,berendsen:1995} utility 
tools, and in-house software developed using GROMACS API.

@@

\begin{figure}[p]
    \caption[Comparison of component mass densities between the HOPC and POPC systems.]{ 
Comparison of component mass densities between the HOPC and POPC systems. 
Densities are computed by dividing the box dimension along the membrane normal into 2000 slices. For 
clairity of presentation, error bars are indicated for every 10th slice. 
Mass densities of solvent and ions, water, lipid headgroup and backbone, and lipid chains are shown. 
The plot for HOPC shows a peak in the water density inside the headgroup region that is not present in POPC, 
similar to what was seen in our previous work with DHPC~\cite{kruczek:2017:ether}. 
}
\label{fig:massdens}
\includegraphics[width=	0.85\textwidth]{dens_3component.eps}
\end{figure}
We find that the headgroup region of 
HOPC has a higher density of solvent compared to POPC -- this is visible as 
a distinct peak in the solvent density profile in the HOPC system. 
We verified that this additional peak does not
result from accumulation of ions by comparing mass density profiles of only water molecules. This
peak is similar to observations of a smaller peak 
made in our previous simulations in the absence of salt,
and may be a direct result of the ether-linked lipids~\cite{kruczek:2017:ether}.
